Eco-evolutionary dynamics of finite populations from first principles

### 2023-04-04 ind_files/99_Appendix_D_examples.tex
Rephrase after proofreading
Add an example for noise-induced selection in m=2 case

\section{An example in one dimension: The stochastic logistic equation}
Here, we analyze example \ref{ex_1D_stoch_logistic}. To recap, we had a population of individuals that exhibit a constant per-capita birth rate $\lambda > 0$, and a per-capita death rate that had the linear density-dependence $\mu + (\lambda-\mu)\frac{n}{K}$, where $\mu$ and $K$ are positive constants. Thus, we have the birth and death rates
\begin{equation}
\label{ex_1D_stoch_logistic_BD_eqns}
\begin{aligned}
    b(n) &= \lambda n\\
    d(n) &= \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n
\end{aligned}
\end{equation}
Here, $K$ is the system-size parameter. Introducing the new variable $x=n/K$, we obtain
\begin{align*}
    b_K(x) &= \frac{1}{K}b(n) = \frac{1}{K}\lambda Kx\\
    d_K(x) &= \frac{1}{K}d(n) = \frac{1}{K}\left(\mu + (\lambda-\mu)\frac{Kx}{K}\right)Kx
\end{align*}
Thus, we have
\begin{equation*}
    A^{\pm}(x) = b_K(x)\pm d_K(x) = x\left(\lambda \pm \left(\left(\mu + (\lambda-\mu)x\right)\right) \right)
\end{equation*}
Defining $r=\lambda-\mu$ and $v=\lambda+\mu$ and using equation \eqref{1D_SDE}, we 
see that the `mesoscopic view' of the system is given by the solution of the SDE
\begin{equation}\label{ex_1D_stoch_logistic_full_SDE}
dX_t =  rX_t(1-X_t)dt + \sqrt{\frac{X_t(v+rX_t)}{K}}dW_t
\end{equation}
From equation \eqref{1D_det_limit}, we see that the deterministic dynamics are
\begin{equation}\label{ex_1D_stoch_logistic_det_limit}
\frac{dx}{dt} = A^-(x) = rx(1-x)
\end{equation}
showing that in the infinite population limit, we obtain the logistic equation. This derivation also makes it clear that two systems with very different stochastic dynamics can nevertheless converge to the same infinite population limit \eqref{ex_1D_stoch_logistic_det_limit}, since equation \eqref{ex_1D_stoch_logistic_det_limit} only depends on the difference $\lambda-\mu$.
\myfig{0.6}{figures/App_stoch_logistic_same_w_different_t.png}{\textbf{Comparison of time series produced by Gillespie simulations of the stochastic logistic equation for two different parameter values.} Simulations with the parameter values $\lambda = 4, \mu = 2$ are plotted in red, and simulations with the parameter values $\lambda = 31, \mu = 29$ are plotted in blue. Both simulations have $K=200$. The infinite population limit of both simulations is plotted in black dotted lines. The graph shows 10 realizations each for the two sets of parameter values.}{fig_1D_same_w_different_t_comparison}
To illustrate the effects of this seemingly innocent fact, figure \ref{fig_1D_same_w_different_t_comparison} compares two simulations which have the same value of $\lambda - \mu$ but a ten-fold difference in $\lambda+\mu$. As is clear from the figure, though both populations have the same behavior at the infinite population limit, populations with a higher value of $\lambda+\mu$ exhibit much wilder fluctuations and are therefore more prone to stochastic extinction. This is the root cause of the noise-induced selection that occurs in higher dimensions, discussed in detail in part \ref{part_summary}.

@@

>>>>

\section{An example in one dimension: The stochastic logistic equation}
Here, we analyze example \ref{ex_1D_stoch_logistic}. To recap, we had a population of individuals that exhibit a constant per-capita birth rate $\lambda > 0$, and a per-capita death rate that had the linear density-dependence $\mu + (\lambda-\mu)\frac{n}{K}$, where $\mu$ and $K$ are positive constants. Thus, we have the birth and death rates
\begin{equation}
\label{ex_1D_stoch_logistic_BD_eqns}
\begin{aligned}
    b(n) &= \lambda n\\
    d(n) &= \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n
\end{aligned}
\end{equation}
Here, $K$ is the system-size parameter. Introducing the population density $x\coloneqq n/K$, we obtain
\begin{align*}
    b_K(x) &= \frac{1}{K}b(n) = \frac{1}{K}\lambda Kx\\
    d_K(x) &= \frac{1}{K}d(n) = \frac{1}{K}\left(\mu + (\lambda-\mu)\frac{Kx}{K}\right)Kx
\end{align*}
Thus, we have
\begin{equation*}
    A^{\pm}(x) = b_K(x)\pm d_K(x) = x\left(\lambda \pm \left(\left(\mu + (\lambda-\mu)x\right)\right) \right)
\end{equation*}
Defining $r=\lambda-\mu$ and $v=\lambda+\mu$ and using equation \eqref{1D_SDE}, we 
see that the `mesoscopic view' of the system is given by the solution of the SDE
\begin{equation}\label{ex_1D_stoch_logistic_full_SDE}
dX_t =  rX_t(1-X_t)dt + \sqrt{\frac{X_t(v+rX_t)}{K}}dW_t
\end{equation}
From equation \eqref{1D_det_limit}, we see that the deterministic dynamics are
\begin{equation}\label{ex_1D_stoch_logistic_det_limit}
\frac{dx}{dt} = A^-(x) = rx(1-x)
\end{equation}
showing that in the infinite population limit, we obtain the logistic equation. This derivation also makes it clear that two systems with very different stochastic dynamics can nevertheless converge to the same infinite population limit \eqref{ex_1D_stoch_logistic_det_limit}, since equation \eqref{ex_1D_stoch_logistic_det_limit} only depends on the difference $\lambda-\mu$.
\myfig{0.6}{figures/App_stoch_logistic_same_w_different_t.png}{\textbf{Comparison of time series produced by Gillespie simulations of the stochastic logistic equation for two different parameter values.} Simulations with the parameter values $\lambda = 4, \mu = 2$ are plotted in red, and simulations with the parameter values $\lambda = 31, \mu = 29$ are plotted in blue. Both simulations have $K=200$. The infinite population limit of both simulations is plotted in black dotted lines. The graph shows 10 realizations each for the two sets of parameter values.}{fig_1D_same_w_different_t_comparison}
To illustrate the effects of this seemingly innocent fact, figure \ref{fig_1D_same_w_different_t_comparison} compares two simulations which have the same value of $\lambda - \mu$ but a ten-fold difference in $\lambda+\mu$. As is clear from the figure, though both populations have the same behavior at the infinite population limit, populations with a higher value of $\lambda+\mu$ exhibit much wilder fluctuations and are therefore more prone to stochastic extinction. This is the root cause of the noise-induced selection that occurs in higher dimensions, discussed in detail in part \ref{part_summary}.

@@

\section{An example of noise-induced selection causing deviations from neutrality despite equal fitness}

To illustrate the biasing effects of noise-induced selection in otherwise neutral dynamics, I will use a simple 2 species Lotka-Volterra competition-like model where the effects of competition are on birth rates of one species but on the death rates of the other.

To motivate this, consider a community that contains two types of birds, say type 1 and type  2. These birds compete for limited resources, but in a peculiar manner: Though the two birds feed on different food sources, the trees that type 1 birds use for nesting are the same as those that the type 2 birds rely on for food. Both types are fiercely territorial and do not tolerate other individuals of either type on either their nesting or feeding sites. Thus, competition between the two types affects the \emph{birth rate} of type 1 birds (because they can't find good nesting sites) but the \emph{death rate} of type 2 birds (because of starvation), whereas intratype competition affects the death rate in both cases (due to competition for food sources). Occasionally, each type can give birth to babies of the other type due to mutations. Let us construct the simplest possible model for such a system.

Let each type of bird have a constant per-capita intrinsic birth rate (rate of birth of individuals, not rate at which individuals give birth) of 1 due to reproduction. Additionally, type 1 birds face a reduction in birth rates due to competition with type 2 birds. Let us assume that the magnitude of this competition (per-capita) is equal to the per-capita competition experienced from other individuals of the same type. Both types have some additional birth rate due to rare mutations of the other type, parameterized by a mutation rate $\mu > 0$. Let $n_i$ be the number of type $i$ individuals (which may vary over time). Assuming trees and birds are both randomly distributed through the landscape, we arrive at the birth rates
\begin{equation}
\label{App_example_stoch_LV_birth_rates}
\begin{aligned}
	b_{1}({n_1},{n_2}) &= {n_1} + \mu n_2 - \frac{{n_1}{n_2}}{K}\\
	b_{2}({n_1},{n_2}) &= {n_2} + \mu n_1
\end{aligned}
\end{equation}
Here, the ${n_1}{n_2}/K$ term represents the effect of competition between types; The product ${n_1}{n_2}$ quantifies how often a type 1 bird and a type 2 bird are expected to interact, and $K$ is a carrying capacity for the habitat (in analogy to logistic growth or Lotka-Volterra competition), and can be thought of as a proxy for the amount of tree cover in the landscape.

For the death rates, I assume that the effect of intra-type competition on the death rate is linearly density-dependent, and thus arrive at the equations:
\begin{equation}
\label{App_example_stoch_LV_death_rates}
\begin{aligned}
	d_{1}({n_1},{n_2}) &= \frac{{n_1^2}}{K}\\
	d_{2}({n_1},{n_2}) &= \frac{{n_2^2}}{K} + \frac{{n_1}{n_2}}{K}
\end{aligned}
\end{equation}
Note that the effect of competition between types manifests here in an increased death rate of type 2 birds due to starvation.

Moving to density space via the change of variables $x_i = n_i/K$, letting $\mathbf{x} = [x_1, x_2]^{\mathrm{T}}$, and comparing terms with equations \eqref{nD_functional_forms_for_replicator}, we can see that the per-capita fitness $w_i$ of each type is:
\begin{equation*}
	w_{1}(\mathbf{x}) = w_{2}(\mathbf{x}) = 1 - x_1 - x_2
\end{equation*}
The two types of birds have the same fitness! This implies that $w_1 = w_2 = \overline{w}$ and the selection term in \eqref{nD_replicator_mutator} vanishes.
However, if we now compute the per-capita turnover rates $\tau_i$  of each type, we see that we have
\begin{align*}
	\tau_{1}(\mathbf{x}) &= 1 + x_1 - x_2\\
	\tau_{2}(\mathbf{x}) &= 1 + x_1 + x_2
\end{align*}
Thus, $\tau_{1} < \tau_{2}$, and from equation \eqref{nD_stochastic_RM}, we thus know that noise-induced selection favors type 1 over type 2.
\myfig{1}{figures/App_stoch_LV_density_plots.png}{\textbf{(A)} Time series and \textbf{(B)} Density estimates for $p$, the fraction of type 1 individuals in the population for various values of $K$, obtained from a direct individual-based simulation of the model  conducted via the Gillespie algorithm. At high $K$, the population conforms to deterministic (infinite population) predictions. The time series are from single realizations. The density plots in panel $(B)$ are estimated from 100 independent realizations, each of which were run for $10^4$ timesteps. All simulations were initialized with $n_1 = n_2 = K/2$. In all simulations, $\mu = 0.05$.}{App_stoch_LV_dens_plots}
Direct Gillespie simulations indeed reveal that for low values of $K$, the fraction of individuals in the population that are of type \textcolor{red}{1} is significantly biased to be greater than 0.5, showing the effect of noise-induced selection (Figure \ref{App_stoch_LV_dens_plots}). This bias disappears for high values of $K$, as expected.

For this model, we can in fact quantitatively derive the effects of noise-induced selection by explicitly calculating each term of equation \eqref{nD_stochastic_RM}. Let $p = x_1/(x_1 + x_2)$ be the frequency of type 1 individuals and let $q = 1-p$. Then, it is easy to check by direct substitution of our functional forms that \eqref{nD_stochastic_RM} becomes
\begin{equation}
	dp =  \frac{2}{K}p^2qdt + \mu\left(1-\frac{1}{KN_K}\right)\left(q-p\right) + q\sqrt{A^{+}_1}dW_t^{(1)} - p\sqrt{A^{+}_2}dW_t^{(2)}
\end{equation}
where $A_i^{+}$ is as defined throughout this thesis and each $W^{i}$ is an independent Wiener process. This equation clearly shows the biasing effect of noise-induced selection in the first term of the RHS. Since $p^2(1-p) > 0$ for $p \in (0,1)$, this term always tends to increase the fraction of type 1 individuals in the population. The second term is a mutational term, and simply reflects the fact that we assumed that $1 \to 2$ and $2 \to 1$ mutations are equally likely (and thus the net direction of mutational effects on frequency depends on the difference between the frequencies of type 1 and type 2 individuals).
